\Wxxsir methods to estimate $\rr$ in the SIR model

### 2018-07-03 paper/Version14/draft_v14.tex
finally on git

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught'') is an elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  In some ways,$\rr$ summarizes an entire outbreak of a disease; it  is used to assess whether a disease epidemic will occur.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$ \citep{anderson1992}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that the estimate $\hat{\rr}$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009}.

To elaborate on $\rr$ being a property of the model, we first need to introduce the concept of the ``SI-framework,'' which was pioneered by Kermack and McKendrick in the 1920s \citep{getz2006}.   Here, S and I are known as compartments where S stands for ``susceptible'' and I for ``infectious.'' Infectious disease models then specify how individuals move from the S compartment to the I compartment.  Often, ancillary compartments are added such as R, which stands for recovered or removed, or E, which stands for exposed but not yet infectious.  The estimate of $\rr$ from a SIR model will not be directly comparable to the estimate of $\rr$ from a SEIR model, for example.  The difficulties of estimating $\rr$ are summarized and expanded upon by \cite{li2011}.  \cite{driessche2017} overviews a number of methods used to estimate $\rr$ but looks at systems beyond the SIR model and so these methods may not be directly compared.   As a result of $\rr$ being a property of the model, we limit our review to methods for estimating $\rr$ for the SIR model introduced by \cite{Kermack700}.

@@

We review \wxxsir methods for estimating $\rr$.  We note that these methods are not exhaustive but are chosen due to their impact and use in epidemiology.  We find that these methods also help to highlight the problems mentioned above.

@@

Other difficulties in estimating $\rr$ arise from model assumptions on how randomness enters the model and sensitivity of the estimate to changes in the model parameters.  For example, the SIR model is special in that two of its compartments, S and R are monotonic.  The number of susceptibles is non-increasing and the number of recovereds is non-decreasing.  Logically, models should enforce this monotonicity of the two compartments.  As a consequence, this restricts assumptions on how randomness enters the model.  For example, Gaussian error, a common assumption in statistics and machine learning, becomes highly suspect due to the assymmetrical nature of the noise imposed by the monotonic $S$ and $R$ compartments.  Without assuming Gaussian noise, it becomes much more difficult to guarantee properties such as unbiased and consistent estimates of $\rr$.

@@

Below in Table \ref{tab:r0-real-ex}, we provide examples of estimates of $\rr$ for specific diseases from publications over the past three decades, spanning various locations over the world.  We display estimates for HIV, Zika (ZIK-V), Ebola (EVD), seasonal influenza, H1N1 influenza, and measles.  These estimates of $\rr$ are not limited to those of SIR model, although most results come from this framework.  Of these, Zika seems to have the largest estimated values of $\rr$, followed by Ebola, H1N1 influenza, seasonal influenza, and HIV.  We see that African locations seem to have larger estimates of $\rr$ compared to the rest of the world, regardless of the disease.  Some of the accompanying intervals (not necessarily CIs) are extremely large such as the case of HIV in Uganda (Interval: [0.12, 14.17]) whereas some are extremely small such as for Ebola in Guinea (Interval: [1.50, 1.52]).  We see that estimates for H1N1 influenza are generally higher than that of seasonal influenza, which is evidence that H1N1 should have been treated as a more critical concern than the typical seasonal influenza cycle.  The lowest possible estimate for $\rr$ reported is 0.12 for HIV in Uganda, and the largest is 17.00 for measles in England and Wales.  Within diseases, we see that estimates for $\rr$ seem to be similar, although the respective CIs typically do not intersect, suggesting that there is disparity in estimates of $\rr$.

@@

In this particular  SIR model (displayed in Figure \ref{fig::sir}), the number of individuals ($N$) is constant. Adaptations of the SIR model can include birth and death rates, which may correspondingly change the derivation of $\rr$ (for further discussion, see Section \ref{sec:discussion}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the inverse of expected time to infection rate and $\gamma$ is the inverse of the expected time to recovery.  The movement of individuals from one compartment to another is represented through the ordinary differential equations below.  Both $\beta$ and $\gamma$ are constrained to be between $0$ and $1$.  For the remainder of this paper, we use $X$, $Y$, and $Z$ to denote the number of individuals in the $S$, $I$, and $R$ compartments, respectively, to avoid any confusion with $\rr$.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Y}. \label{eq:sir}
\end{align}
In words, susceptible individuals become infected at a rate that is proportional to the percentage of infected individuals multiplied by $\beta$, the inverse of expected time to infection, and the number of susceptible individuals.  Infectious individuals recover at a rate of $\gamma$ multiplied by the number of infected individuals.

@@

\end{tikzpicture}
\caption{Depiction of a SIR model where $X=S, Y=I,$ and $Z=R$.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}
An outbreak occurs if the rate of change of infectious individuals is positive, $\frac{dY}{dt} > 0$, or equivalently,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
  \implies -\frac{dX}{dt} - \frac{dZ}{dt} &> 0 \\
  \implies \frac{\beta X Y}{N}  - \gamma Y &> 0 ,\\
\implies  Y \left ( \beta \frac{X}{N} - \gamma \right ) & > 0\\
\implies   \frac{\beta}{\gamma} &> \frac{N}{X}.
\end{align*}
That is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of initially susceptible individuals is large compared to the total population, $\frac{N}{X} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align*}
  \rr \overset{def}{=} \frac{\beta}{\gamma}.
  \end{align*}
  To incorporate randomness into the model, we add noise, namely,
  \begin{align}\label{eq:sir-noise}
    X_{obs}(t) &= X(t) + \epsilon_{X,t}\\
    Y_{obs}(t) &=  N - X_{obs}(t) -Y_{obs}(t)  \nonumber\\
    Z_{obs}(t) &= Z(t) - \epsilon_{Z,t}. \nonumber
  \end{align}
We are assuming the observations are generated based on the discretized time (replacing $d$ with $\Delta$), deterministic ODEs presented in Equation \ref{eq:sir} with the addition of time ($t=0, 1, \dots, T$) and compartment dependent noise $\epsilon_{X,t}$ and $\epsilon_{Z,t}$.  Since, $N$, the total population, is constant, then $Y_{obs}$ is adjusted accordingly.

@@

Good properties of an estimator may include (1) being a consistent estimator
\begin{align*}
  \hat{\rr} \overset{P}{\to} \rr \textnormal{ as } T\to \infty,
\end{align*}
where `$\overset{P}{\to}$' stands for convergence in probability \citep{wasserman2004} or (2) converging to a known probability distribution (convergence in distribution `$\overset{d}{\to}$' \citep{wasserman2004}),
\begin{align*}
\hat{\rr} \overset{d}{\to} F.  
\end{align*}
As we are interested in the hypothesis
\begin{align*}
  H_0:\;& \rr > 1 \\
  H_A:\;& \rr \le 1
\end{align*}
Then for some $\alpha \in [0,1]$, we would like to know for some $\hat{a}, \hat{b} >0$
\begin{align*}
P(\hat{a} \le \rr \le \hat{b}) \le 1 - \alpha.
\end{align*}
This property is particularly desirable, as we can estimate $\hat{a}$ and $\hat{b}$ from the distribution $F$.  At any rate, we would like to be able to estimate $\hat{a}$ and $\hat{b}$ for a given $\alpha$-level.  We could also make similar hypothesis tests such as that the $\rr$ for H1N1 influenza is greater than the reported $\rr$ for seasonal influenza.

\subsection{$\rr$ is a property of the model}\label{sec:prop}

@@

Consider the case when $N=10000, (\beta, \gamma) = (.06, .03)$, $(X(0), E(0), Y(0), Z(t)) = (9500, 0, 500, 0)$ for times $t=1, \dots, 365$.  Let the parameter $\alpha$ range between [0.005, 0.25] with step size $0.005$.  This generates a first data set
$$D_{\alpha, \textnormal{XEYZ}} = \left \{(X_{XEYZ}(t), E_{XEYZ}(t), Y_{XEYZ}(t), Z_{XEYZ}(t): t=1, \dots, 365\right \},$$
which is observed by researcher 1.  Now assume that researcher 2 misspecifies the model for this data as a XYZ model rather than a XEYZ model.  That is, she observes all exposed individuals as susceptible and generates a second data set
$$D_{\alpha, XYZ} = \left \{X_{XYZ}(t), Y_{XYZ}(t), Z_{XEYZ}(t) : t = 1, \dots, 365 \right \},$$
where $X_{XYZ}(t) = X_{XEYZ}(t) + E_{XEYZ}(t)$, $Y_{XYZ}(t) = Y_{XEYZ}(t)$, and  $Z_{XEYZ}(t) = Z_{XYZ}(t)$. This situation is illustrated in Figure \ref{fig:sir-vs-seir} with $\alpha = 0.01$.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{images/r0-est.pdf}
  \caption{Estimates of $\rr$ from both the XEYZ and XYZ model specification for different values of $\alpha$ using Least Squares with $\beta=0.06$, $\gamma=0.03$, and $(X(0), E(0), Y(0)) = (9500, 500, 0)$.}\label{fig:r0-alpha}
  \end{figure}

For those readers not convinced by our small example, we offer this mathematical intuition of the next generation model (NGM) described by \cite{diekmann1990} and \cite{diekmann2009}.

 The NGM is a generalization of any compartment model at the infection-free steady state. This model solves the problem of having an expression for $\rr$ in terms of the epidemiological parameters for a general compartment model.  Originally, introduced by \citep{diekmann1990}, \cite{diekmann2009} posit a recipe to find $\rr$ for a wide class of compartment models, including SIR and SEIR. First, define the infected subsystem as ``the equations of the ODE system that describe the production of new infections and changes in state among infected individuals.''  Let $x = (C_1, C_2, \dots, C_m)$ where $C_i$ are the different compartments of the infected subsystem.  The steps to find $\rr$ are as follows:

\begin{enumerate}
\item ``Linearize the infected subsystem of nonlinear ODEs about the infection-free steady state''
\item Decompose the linearized infected subsystem into $(T + \Sigma )x$ where $T$ is a $m\times m$ matrix of transmissions and $\Sigma$ is a $m \times m$ matrix of transitions.
\item $\rr$ is the spectral radius (i.e. the dominant eigenvalue) of $-T \Sigma^{-1}$.  
\end{enumerate}

Here, $T_{ij}$ is the rate of transmission of \textit{newly} infected individuals in state $i$ created by individuals in state $j$.  $\Sigma_{ij}$ is the transition rate of individuals into compartment $i$ from compartment $j$.

This method has the assumptions of a basic compartment model: homogeneous mixing among populations and the law of mass action, i.e., that the number of infected individuals are proportional to the number of infected individuals at the previous step.   Additionally, this method is advantageous in that the matrix $-\Sigma^{-1}$ has intuitive meaning.  The $ij$th entry of $- \Sigma^{-1}$ is the ``expected time that an individual who presently has state $j$ will spend in infected state $i$.''

The NGM shows that $\rr$ is the spectral radius of a matrix composed from the transmission of disease from diseased individuals to susceptible individuals and transition of diseased individuals from among the compartments and is indeed a property of the model.

@@

In Section \ref{sec:prop} we described how $\rr$ is a property of the model.  To fairly compare estimation methods to one another, we restrict our methods to estimate $\rr$ to the SIR model.  Also in Section \ref{sec:prop}, we examined simulated data that was generated from a SEIR model without noise.  However, we know that data is very noisy.    In this section, we describe the \wxxsir methods to estimate $\rr$ for the SIR model in the presence of noise in detail, including advantages and disadvantages of each.

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the least squares (LS) estimate for $\hat{\beta}$ and $\hat{\gamma}$ is then combined to give an estimate of $\hat{\rr}$ in Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

@@

\begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dX}{dt} &= - \rr \gamma Y \frac{X}{N}\vspace{.5em}\\
    \frac{dY}{dt} &=  \rr \gamma Y \frac{X}{N}  - \gamma Y\vspace{.5em} \\
    \frac{dZ}{dt} &=  - \gamma Y 
  \end{array}
  \right . .
  \end{align*}
We find
\begin{align}
(\hat{\rr}, \hat{\gamma} ) &= \text{argmin}_{\rr, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align}
We use the $\hat{\rr}$ directly from the above estimation problem, which again can be found with a grid search or another optimization process.

We can also write $\hat{\rr}$ as a function of $\gamma^*$,
\begin{align*}
\hat{\rr}(\gamma^*)&= \text{argmin}_{\rr} \sum_{t} \left [ \left (X_{obs}(t) - X(t; \gamma^*)\right )^2 + \left ( Z_{obs}(t) - Z(t; \gamma^*) \right )^2 \right ].
  \end{align*}

It may be surprising that this method leads to different results than simply using least squares.  We also see that this method further results in different standard errors than in LS, which we discuss further in Section \ref{sec:results}.

@@

Specifically, we estimate two polynomials in \(t\) with degree $K$  to \(X_{obs}\)
and \(Z_{obs}\) using least squares to find the coefficients $\{(\hat{x}_k,
\hat{z}_k)\}_{k=1, \dots, K}$,
\begin{align*}
\hat{X}(t) &= \sum_{k=0}^K \hat{x}_k t^k\\
{\hat{Z}}(t) &= \sum_{k=0}^K \hat{z}_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Z}^\prime(t) &= \sum_{k=0}^K k \hat{z}_k t^{k-1}
\end{align*}
An estimator for $\rr$ is derived from the ODEs in Equation \eqref{eq:sir},
\begin{align}
  - \frac{X^\prime}{Z^\prime}&= \rr \frac{X}{N} \nonumber\\
  \rr &=       -\frac{X^\prime}{
        Z^\prime} \cdot \frac{N}{X} \nonumber\\
  \hat{\rr} &= -\frac{\hat{X}^\prime(0)}{ \hat{Z}^\prime(0)} \cdot \frac{N}{\hat{X}(0)}. \nonumber
  \end{align}
  Here, $K$ is arbitrary and should be selected using some criterion such as AIC (REF).  Besides optionally deciding on the degree of polynomials to fit, this model is simple to implement and gives comparable results to using least squares with the SIR model.  The time $t=0$ is used to best capture the initial outbreak.

@@

The above formulation (Section \ref{linear-model-approximation-degree-10}) of a linear model approximation only uses the estimate at time $t=0$ to estimate $\rr$.  We can instead, use all time points $T$ available to estimate $\rr$.  We fit a polynomial in \(t\) with degree \(K\) to \(X_{obs}\)
and \(Z_{obs}\) as above, with a slight modification in how we estimate
\(\rr\),
\begin{align*}
  \hat{\rr} &= \frac{1}{T} \sum_t \frac{-\hat{X}^\prime(t)}{\hat{Z}^\prime(t)} \cdot \frac{N}{X(0)} 
\end{align*}
The intuition is that $\frac{-X^\prime(t)}{Z^\prime(t)}$ is constant in $t$, but due to our approximations with the linear model, this is no longer the case.  Here, we average over the different possible values of $\rr$, estimated at different times.  An advantage to this approach is that we have a more robust estimate of $\rr$ than just using one time point.  As in LMA, we can examine how sensitive $\rr$ is to the order of the polynomial $K$.

@@

The advantages of this method are many.  One, the estimate of $\rr$ is highly interpretable.  Two, an assumption of Gaussian noise in Equation \ref{eq:harko_lin} is not as strong as an assumption in the other methods to estimate $\rr$ because we are looking at a log transformation of the percent of suceptibles instead of raw variables.  If the assumption holds, we can show that the resulting estimate of $\rr$ is unbiased and consistent.  Finally,  this method is easily implemented using any linear modeling software.

@@

The advantages of this model are its simplicity of interpretation and ability to generate an estimate for the probability distrubtion for $(X(t), Y(t))$.  We have explicitly stated how randomness enters the model so it is up to the researcher to check whether this assumption is a good fit.  Using this method, it is easy to see how sensitive $\hat{\rr}$ is to changes in $\hat{\alpha}$ as $\hat{\rr}$ is a function of $\alpha$.

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SIR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1) | \rr  \sim \textnormal{Poisson}( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous time period's incidence is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous time period.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

A choice one must make in the block bootstrap is the block size $b$.  One simple recommendation is to take $b= Cn^{1/3}$, which is derived from the asymptotic mean square error of the method.  \cite{hall1995} describe how to calculate $C$, which is context dependent.

>>>>

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught'') is an elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  In some ways,$\rr$ summarizes an entire outbreak of a disease; it  is used to assess whether a disease epidemic will occur.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$ \citep{anderson1992}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009}.

To elaborate on $\rr$ being a property of the model, we first need to introduce the concept of the ``SI-framework,'' which was pioneered by Kermack and McKendrick in the 1920s \citep{getz2006}.   Here, S and I are known as compartments where S stands for ``susceptible'' and I for ``infectious.'' Infectious disease models then specify how individuals move from the S compartment to the I compartment.  Often, ancillary compartments are added such as R, which stands for recovered or dead, or E, which stands for exposed but not yet infectious.  The estimate of $\rr$ from a SIR model will not be directly comparable to the estimate of $\rr$ from a SEIR model, for example.  The difficulties of estimating $\rr$ are summarized and expanded upon by \cite{li2011}.  \cite{driessche2017} overviews a number of methods used to estimate $\rr$ but looks at systems beyond the SIR model and so these methods may not be directly compared.   As a result of $\rr$ being a property of the model, we limit our review to methods for estimating $\rr$ for the SIR model introduced by \cite{Kermack700}.

@@

We review \wxxsir methods for estimating $\rr$.  We note that these methods are not exhaustive but are chosen due to their impact and use in epidemiology.  We find that these methods also help to highlight the problems discussed above.

@@

Other difficulties in estimating $\rr$ arise from model assumptions on how randomness enters the model and sensitivity of the estimate to changes in the model parameters.  For example, the SIR model is special in that two of its compartments, S and R are monotonic.  The number of susceptibles is non-increasing and the number of susceptibles is non-decreasing.  Logically, models should enforce this monotonicity of the two compartments.  As a consequence, this restricts assumptions on how randomness enters the model.  For example, Gaussian error, a common assumption in statistics and machine learning, becomes highly suspect due to the non-symmetrical nature of the noise imposed by the monotonic $S$ and $R$ compartments.  Without assuming Gaussian noise, it becomes much more difficult to guarantee properties such as unbiased and consistent estimates of $\rr$.

@@

Below in Table \ref{tab:r0-real-ex}, we provide examples of estimates of $\rr$ for specific diseases from publications over the past three decades, spanning various locations over the world.  We display estimates for HIV, Zika (ZIK-V), Ebola (EVD), seasonal influenza, H1N1 influenza, and measles.  These estimates of $\rr$ are not limited to those of SIR model, although most results come from this framework.  Of these, Zika seems to have the largest estimated values of $\rr$, followed by Ebola, H1N1 influenza, seasonal influenza, and HIV.  We see that African locations seem to have larger estimates of $\rr$ compared to the rest of the world, regardless of the disease.  Some of the accompanying intervals (not necessarily CIs) are extremely large such as the case of HIV in Uganda (Interval: [0.12, 14.17]) whereas some are extremely small such as for Ebola in Guinea (Interval: [1.50, 1.52]).  We see that estimates for H1N1 influenza are generally higher than that of seasonal influenza, which is evidence that H1N1 should have been treated as a more critical concern than the typical seasonal influenza cycle.  The lowest possible estimate for $\rr$ reported is 0.12 for HIV in Uganda, and the largest is 17.00 for measles in England and Wales.  Within diseases, we see that estimates for $\rr$ seem to be similar, although typically do not lie in one another's intervals, suggesting that there is disparity in estimates of $\rr$.

@@

In this particular  SIR model (displayed in Figure \ref{fig::sir}), the number of individuals ($N$) is constant. Adaptations of the SIR model can include birth and death rates, which may correspondingly change the derivation of $\rr$ (for further discussion, see Section \ref{sec:discussion}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the average infection rate and $\gamma$ is the average recovery rate.  The movement of individuals from one compartment to another is represented through the ordinary differential equations below.  Both $\beta$ and $\gamma$ are constrained to be between $0$ and $1$.  For the remainder of this paper, we use $X$, $Y$, and $Z$ to denote the number of individuals in the $S$, $I$, and $R$ compartments, respectively, to avoid any confusion with $\rr$.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Y}. \label{eq:sir}
\end{align}
In words, susceptible individuals become infected at a rate that is proportional to the percentage of infected individuals multiplied by $\beta$, the infection rate, and the number of susceptible individuals.  Infectious individuals recover at a rate of $\gamma$ multiplied by the number of infected individuals.

@@

\end{tikzpicture}
\caption{Depiction of a SIR model where $X=S, Y=I,$ and $Z=R$.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}
An outbreak occurs if the rate of change of infectious individuals is positive, $\frac{dY}{dt} > 0$, or equivalently,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
  \implies -\frac{dX}{dt} - \frac{dZ}{dt} &> 0 \\
  \implies \frac{\beta X Y}{N}  - \gamma Y &> 0 ,\\
\implies  Y \left ( \beta \frac{X}{N} - \gamma \right ) & > 0\\
\implies   \frac{\beta}{\gamma} &> \frac{N}{X}.
\end{align*}
That is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of initially susceptible individuals is large compared to the total population, $\frac{X}{N} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align*}
  \rr \overset{def}{=} \frac{\beta}{\gamma}.
  \end{align*}
  To incorporate randomness into the model, we add noise, namely,
  \begin{align}\label{eq:sir-noise}
    X_{obs}(t) &= X(t) + \epsilon_{X,t}\\
    Y_{obs}(t) &=  N - X_{obs}(t) -Y_{obs}(t)  \nonumber\\
    Z_{obs}(t) &= Z(t) - \epsilon_{Z,t}. \nonumber
  \end{align}
We are assuming the observations are generated based on the discretizied time, deterministic ODEs presented in Equation \ref{eq:sir} with the addition of time ($t=0, 1, \dots, T$) and compartment dependent noise $\epsilon_{X,t}$ and $\epsilon_{Z,t}$.  Since, $N$, the total population, is constant, then $Y_{obs}$ is adjusted accordingly.

@@

Good properties of an estimator may include (1) being a consistent estimator
\begin{align*}
  \hat{\rr} \overset{P}{\to} \rr \textnormal{ as } T\to \infty,
\end{align*}
where `$\overset{P}{\to}$' stands for convergence in probability \citep{wasserman2004} or (2) converging to a known probability distribution (convergence in distribution `$\overset{d}{\to}$' \citep{wasserman2004}),
\begin{align*}
\hat{\rr} \overset{d}{\to} F.  
\end{align*}
As we are interested in the hypothesis
\begin{align*}
  H_0:\;& \rr < 1 \\
  H_A:\;& \rr \ge 1
\end{align*}
Then for some $\alpha \in [0,1]$, we would like to know for some $\hat{a}, \hat{b} >0$
\begin{align*}
P(\hat{a} \le \rr \le \hat{b}) \le 1 - \alpha,  
\end{align*}
the second property is particularly desirable, as we can estimate $\hat{a}$ and $\hat{b}$ from the distribution $F$.  At any rate, we would like to be able to estimate $\hat{a}$ and $\hat{b}$ for a given $\alpha$-level.  We could also make similar hypothesis tests such as that the $\rr$ for H1N1 influenza is greater than the reported $\rr$ for seasonal influenza.

\subsection{$\rr$ is a property of the model}

@@

Consider the case when $N=10000, (\beta, \gamma) = (.06, .03)$, $(X(0), E(0), Y(0), Z(t)) = (9500, 0, 500, 0)$ for times $t=1, \dots, 365$.  Let the parameter $\alpha$ range between [.005, 1].  This generates data set
$$D_{\alpha, \textnormal{XEYZ}} = \left \{(X_{XEYZ}(t), E_{XEYZ}(t), Y_{XEYZ}(t), Z_{XEYZ}(t): t=1, \dots, 365\right \},$$
which is observed by researcher 1.  Now assume that researcher 2 misspecifies the model for this data as a XYZ model rather than a XEYZ model.  That is, she places all exposed individuals in the X compartment to have data set
$$D_{\alpha, XYZ} = \left \{X_{XYZ}(t) = X_{XEYZ} + E_{XEYZ}, Y_{XYZ} = Y_{XEYZ}, Z_{XEYZ} = Z_{XYZ} : t = 1, \dots, 365 \right \}.$$
This situation is illustrated in Figure \ref{fig:sir-vs-seir} with $\alpha = 0.01$.

@@

For those

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{images/r0-est.pdf}
  \caption{Estimates of $\rr$ from both the XEYZ and XYZ model specification for different values of $\alpha$.}\label{fig:r0-alpha}
  \end{figure}

@@

In this section, we describe the \wxxsir methods in detail, including advantages and disadvantages of each.

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the least squares (LS) estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

@@

\begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dX}{dt} &= - \rr \gamma Y \frac{X}{N}\vspace{.5em}\\
    \frac{dY}{dt} &=  \rr \gamma Y \frac{X}{N}  - \gamma Y\vspace{.5em} \\
    \frac{dZ}{dt} &=  - \gamma Y 
  \end{array}
  \right . .
  \end{align*}
We find
\begin{align*}
(\hat{\rr}, \hat{\gamma} ) &= \text{argmin}_{\rr, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
We use the $\hat{\rr}$ directly from the above estimation problem, which again can be found with a grid search or another optimization process.  It may be surprising that this method leads to different results than simply using least squares.  One reason we obtain different results may be attributed to better numerical precision as there is one fewer step to obtain an estimate for $\rr$.  However, we also see that this method further results in different standard errors than in LS, which we discuss further in Section \ref{sec:results}.

@@

Specifically, we estimate two polynomials in \(t\) with degree $K$  to \(X_{obs}\)
and \(Z_{obs}\) using least squares to find the coefficients $\{(\hat{x}_k,
\hat{z}_k)\}_{k=1, \dots, K}$,
\begin{align*}
\hat{X}(t) &= \sum_{k=0}^K \hat{x}_k t^k\\
{\hat{Z}}(t) &= \sum_{k=0}^K \hat{z}_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Z}^\prime(t) &= \sum_{k=0}^K k \hat{z}_k t^{k-1}
\end{align*}
Following,  an estimator for \(\rr\) is derived from the ODEs in Equation \eqref{eq:sir},
\begin{align}
  - \frac{X^\prime}{Z^\prime}&= \rr \frac{X}{N} \nonumber\\
  \rr &=       -\frac{X^\prime}{
        Z^\prime} \cdot \frac{N}{X} \nonumber\\
  \hat{\rr} &= -\frac{\hat{X}^\prime(0)}{ \hat{Z}^\prime(0)} \cdot \frac{N}{\hat{X}(0)}. \nonumber
  \end{align}
  Here, $K$ is arbitrary and should be selected using some criterion such as AIC.  Besides optionally deciding on the degree of polynomials to fit, this model is simple to implement and gives comparable results to using least squares with the SIR model.  The time $t=0$ is used to best capture the initial outbreak.

@@

The above formulation (Section \ref{linear-model-approximation-degree-10}) of a linear model approximation only uses the estimate at time $t=0$ to estimate $\rr$.  We can instead, use all time points $O$ available to estimate $\rr$.  We fit a polynomial in \(t\) with degree \(K\) to \(X_{obs}\)
and \(Z_{obs}\) as above, with a slight modification in how we estimate
\(\rr\),
\begin{align*}
  \hat{\rr} &= \frac{1}{O} \sum_t \frac{-\hat{X}^\prime(t)}{\hat{Z}^\prime(t)} \cdot \frac{N}{X(0)} 
\end{align*}
The intuition is that $\frac{-X^\prime(t)}{Z^\prime(t)}$ is constant in $t$, but due to our approximations with the linear model, this is no longer the case.  Here, we average over the different possible values of $\rr$, estimated at different times.  An advantage to this approach is that we have a more robust estimate of $\rr$ than just using one time point.  Like in LMA, we can examine how sensitive $\rr$ is to the order of the polynomial $K$.

@@

The advantages of this method are many.  One, the estimate of $\rr$ is highly interpretable.  Two, an assumption of Gaussian noise in Equation \ref{eq:harko_lin} is not as strong as an assumption in the other methods to estimate $\rr$ because we are looking at a log transformation of the the percent of suceptibles instead of raw variables.  If the assumption holds, we can show that the resulting estimate of $\rr$ is an unbiased and a consistent estimator.  Finally,  this method is easily implemented using any linear modeling software.

@@

The advantages of this model are its simplicity of interpretation and ability to generate an estimate for the probability distrubtion for $(X(t), Y(t))$.  We have explicitly stated how randomness enters the model so it is up to the researcher to check whether this assumption is a good fit.  Using this method, it is possible to simulate how sensitive $\hat{\rr}$ is to changes in $\hat{\alpha}$.

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SIR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1) | \rr  \sim \textnormal{Poisson}( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidence is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

A choice one must make in the block bootstrap is the block size $b$.  One simple recommendation is to take $b= n^{1/3}$, which is derived from the asymptotic mean square error of the method.
